Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys

### 2019-02-01 doc/article.tex
Helene, Dragan-2 comments in

\begin{document}
%\raggedright
\begin{raggedright}
\huge
Astro2020 Science White Paper \linebreak

Testing Gravity Using Type Ia Supernovae Discovered by Next-Generation Wide-Field Imaging Surveys \linebreak
\normalsize

@@

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SN absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands; over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
 for which good light curves can be measured, corresponding to a  number density of $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$.
  This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.

Given these two advances, supernovae discovered by wide-field searches in the next decade will measure the growth of structure.
For example, 
over the course of a decade a SN survey relying on LSST discoveries 
plus spectroscopic redshifts can produce find 4--14\% uncertainties in $fD$ in $0.05$ redshift bins from $z=0$ to 0.3, cumulatively giving
2.2\% uncertainty on fD  within this interval , where at
$0<z<0.2$ most of the probative power comes from peculiar velocities and at higher redshifts from $RSD$  \citep{2017ApJ...847..128H}.

\section{Testing Gravity with Peculiar Velocity Surveys}
While the growth rate $fD$ can be used to test several aspects of physics beyond the standard cosmological model (e.g.\ dark matter clustering, dark energy evolution), our scientific interest here is in probing gravity
so we here focus  on $\gamma$.
To illustrate the distinction,
 $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega_M + \int \Omega_M^{\gamma} \ln{\Omega_M}\, d\ln{a} \approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ respectively
 in $\Lambda$CDM;  two surveys with the same fractional precision  in $fD$ will have different precision in $\gamma$,
 with the one at lower redshift providing the tighter constraint. 
 In this section, we focus on how well peculiar velocity surveys in the upcoming decade can measure $\gamma$ and the sensitivity of the measurement
 on survey parameter choices.

We project uncertainties in $\gamma$, $\sigma_\gamma$ for a suite of idealized surveys
using a Fisher matrix analysis similar to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}.
The Fisher information matrix is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}
\label{cov:eq}
\end{equation}
and the two parameters considered are $\lambda \in \{\gamma,bD\}$.
The parameter dependence enters through $(fD)(\gamma)$ in the relations $P_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy overdensity
power spectrum $P_{\delta \delta }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.  While the $bD$ term does contain information on $\gamma$, its constraining power is not used here.
The uncertainty in $\gamma$ is $\sigma_\gamma = \sqrt{\left(F^{-1}\right)_{\gamma\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or non-constant $\gamma$, such observations provide additional leverage in probing gravity
but are not considered.

The uncertainty  $\sigma_\gamma$ of a survey depends on its solid angle $\Omega$, depth $r_{max}= \chi(z_{max})$, duration
$t$ through $n=\epsilon \phi t$ where $\phi$ is the observer-frame SN~Ia rate
and $\epsilon$ is the sample-selection efficiency, and
the intrinsic SN~Ia magnitude dispersion 
through $\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$.
An analysis in which the RSD and velocity surveys are treated independently 
sets the off-diagonal elements of $C$ to zero.

@@

@@

{\bf SNe~Ia Need Precise Distances:} Cosmological redshifts are inferred from SN~Ia distances. Distances are calculated from multi-band light curves, redshifts (when available),
SN spectroscopic features, and
host-galaxy properties.
Distance uncertainties here have been characterized by $\sigma_M$,
which includes both measurement uncertainty and intrinsic magnitude dispersion.  The latter depends on the data-quality and methodology used to calculate distances.
For the same supernovae, the intrinsic distance dispersion derived from a dataset with sparse light curves, a small number of bands, limited wavelength coverage, and
lacking spectral-feature measurements is larger compared to the dispersion derived from a broader dataset  with dense light curves, a large number of bands, 
broad wavelength coverage, and with spectral-feature measurements.  Historically surveys with 2 or 3 well-sampled optical light curves obtained $\sigma_M \approx 0.15$~mag,
with dense, $griz$-band optical light curves DES \citep{2018arXiv181102377B} gets $\sigma_M\approx 0.10$~mag, 
standardization studies show that supplemental
infrared data \citep{2012MNRAS.425.1007B} or spectrophotometry at peak brightness
 \citep{2015ApJ...815...58F} are projected to give $\sigma_M \lesssim 0.08$~mag. The improvement between $\sigma_M=0.08$ and 0.15~mag dispersions is equivalent to a factor of 3.52
 in number density, or equivalently in survey duration.

As a point of reference, a practically complete (accounting for temporal observing gaps) discovery up to $z_{max}=0.2$ would have $\sim 150,000$, $r \lesssim 20.5$~mag targets over 10 years.  
Focusing on low-redshift supernovae  mitigates  two important uncertainties
associated with the high-$z$ SN~Ia cosmology. The measurement is relatively insensitive to absolute color calibration, due to the limited
range of {\it observer} wavelengths used to standardize restframe distance moduli.
Depending on which cameras are used to derive distances, instrumental, as opposed to absolute,
calibration may be sufficient.  The progenitor population  is not expected to evolve significantly in the $0.01<z<0.2$ Universe.

@@

%SNe~Ia are already powerful probes of the homogeneous cosmological expansion of the Universe.  
In the next decade,
the high number of SN discoveries together with improved precision in their distance precisions will make $z<0.2$ SNe~Ia, more so than
galaxies,  powerful probes of gravity through their effect  on the growth of structure. 
A peculiar velocity survey must discover, classify, and get distances of SNe~Ia, as well as obtain redshifts of their host galaxies.
Choices in survey design impact the precision obtained on $\gamma$.  SN~Ia peculiar velocity surveys are of the highest scientific
interest and we encourage
the community to develop aggressive surveys in the pursuit of testing  General Relativity and probing gravity.

@@

Each supernova is assigned a magnitude based on the distance modulus of its cosmological
redshift plus a random term drawn from a Normal distribution $\sigma_M$, which for simplicity is the same for all supernovae and captures both
intrinsic magnitude dispersion and measurement uncertainty.  No other corrections are applied to the observed magnitude.  Dipole effects of the heliocentric
motion with respect to the CMB and of the galaxies with respect to the CMB  are ignored: referring to
\citet{2011ApJ...741...67D}, the effects in Eq.~18 are not included.  Including these effects are straightforward but is computationally expensive
and does not affect our conclusions.

>>>>

\begin{document}
%\raggedright
\begin{raggedright}
\huge
Astro2020 Science White Paper \linebreak
proj
Testing Gravity Using Type Ia Supernovae Discovered by Next-Generation Wide-Field Imaging Surveys \linebreak
\normalsize

@@

Two advances in the upcoming decade will make SN~Ia peculiar velocities more powerful.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields  absolute magnitude
dispersion $\sigma_M \gtrsim 0.12$~mag.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SN absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands; over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia
 for which good light curves can be measured, corresponding to a  number density of $n \sim 5\times 10^{-4}h^3$\,Mpc$^{-3}$.
  This sample has comparable
 number density and more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.

Given these two advances, supernovae discovered by wide-field searches in the next decade will 
 be able to stringently constrain the growth of structure in the low-redshift Universe.
For example, 
over the course of a decade a SN survey relying on LSST discoveries 
plus spectroscopic redshifts can produce find 4--14\% uncertainties in $fD$ in $0.05$ redshift bins from $z=0$ to 0.3, cumulatively giving
2.2\% uncertainty on fD  within this interval , where at
$0<z<0.2$ most of the probative power comes from peculiar velocities and at higher redshifts from $RSD$  \citep{2017ApJ...847..128H}.

\section{Testing Gravity with Peculiar Velocity Surveys}
While the growth rate $fD$ can be used to test several aspects of physics beyond the standard cosmological model (e.g.\ dark matter clustering, dark energy evolution), our scientific interest here is in probing gravity
so we here focus  on the growth index $\gamma$.
To illustrate the distinction,
 $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega_M + \int \Omega_M^{\gamma} \ln{\Omega_M}\, d\ln{a} \approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ respectively
 in $\Lambda$CDM;  two surveys with the same fractional precision  in $fD$ will have different precision in $\gamma$,
 with the one at lower redshift providing the tighter constraint. 
 In this section, we focus on how well peculiar velocity surveys in the upcoming decade can measure $\gamma$ and the sensitivity of the measurement
 on survey parameter choices.

We project uncertainties  on the growth index, $\sigma_\gamma$ for a suite of idealized surveys
using a Fisher matrix analysis similar to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}
\citep[there is an alternative approach using an estimator for the mean pairwise velocity][]{2011PhRvD..83d3004B}.
The Fisher information matrix is
\begin{align}
F_{ij} 
%& = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
%\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
where
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{\delta \delta}(k,\mu) + \frac{1}{n} &
   P_{v\delta}(k,\mu)  \\
   P_{v\delta}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}
\label{cov:eq}
\end{equation}
and the two parameters considered are $\lambda \in \{\gamma,bD\}$.
The parameter dependence enters through $(fD)(\gamma)$ in the relations $P_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy overdensity
power spectrum $P_{\delta \delta }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $P_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu$ is the cosine of the angle between the $k$-mode
and the line of sight.  While the $bD$ term does contain information on $\gamma$, its constraining power is not used here.
The uncertainty in $\gamma$ is $\sigma_\gamma = \sqrt{\left(F^{-1}\right)_{\gamma\gamma}}$.
Non-GR models may  also predict a change in the scale-dependence of the growth or non-constant $\gamma$, such observations provide additional leverage in probing gravity
but are not considered.

The uncertainty  $\sigma_\gamma$ of a survey depends on its solid angle $\Omega$, depth
given by the comoving distance out to the maximum redshift $r_{max}=r(z_{max})$, duration
$t$ through $n=\epsilon \phi t$ where $\phi$ is the observer-frame SN~Ia rate
and $\epsilon$ is the sample-selection efficiency, and
the intrinsic SN~Ia magnitude dispersion 
through $\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$.
An analysis in which the RSD and velocity surveys are treated independently 
sets the off-diagonal elements of $C$ to zero.

@@

The dependence on the uncertainty in the growth index on the survey parameters is as follows:

@@

{\bf SNe~Ia Need Precise Distances:} SN~Ia distances, along with their measured redshifts, are required to infer the peculiar velocities of their hosts.
Distances are calculated from multi-band light curves, redshifts (when available),
SN spectroscopic features, and
host-galaxy properties.
Distance uncertainties here have been characterized by magnitude dispersion $\sigma_M$,
which includes both measurement uncertainty and intrinsic magnitude dispersion.  The latter depends on the data-quality and methodology used to calculate distances.
For the same supernovae, the intrinsic distance dispersion derived from a dataset with sparse light curves, a small number of bands, limited wavelength coverage, and
lacking spectral-feature measurements is larger compared to the dispersion derived from a broader dataset  with dense light curves, a large number of bands, 
broad wavelength coverage, and with spectral-feature measurements.  Historically surveys with two or three well-sampled optical light curves obtained $\sigma_M \approx 0.15$~mag,
with dense, $griz$-band optical light curves DES \citep{2018arXiv181102377B} gets $\sigma_M\approx 0.10$~mag, 
standardization studies show that supplemental
infrared data \citep{2012MNRAS.425.1007B} or spectrophotometry at peak brightness
 \citep{2015ApJ...815...58F} are projected to give $\sigma_M \lesssim 0.08$~mag. The improvement between $\sigma_M=0.08$ and 0.15~mag dispersions is equivalent to a factor of 3.52
 in number density, or equivalently in survey duration.

As a point of reference, a practically complete (accounting for temporal observing gaps) discovery up to $z_{max}=0.2$ would have $\sim 150,000$, $r \lesssim 20.5$~mag targets over 10 years.  
Focusing on low-redshift supernovae  mitigates  several important uncertainties
associated with the high-$z$ SN~Ia cosmology. The measurement is relatively insensitive to absolute color calibration, due to the limited
range of {\it observer} wavelengths used to standardize restframe distance moduli.
Depending on which cameras are used to derive distances, instrumental, as opposed to absolute,
calibration may be sufficient.  The progenitor population  is not expected to evolve significantly in the $0.01<z<0.2$ Universe.

@@

%SNe~Ia are already powerful probes of the homogeneous cosmological expansion of the Universe.  
In the next decade,
the high number of SN discoveries together with improved precision in their distance precisions will make $z<0.2$ SNe~Ia, more so than
galaxies,  powerful probes of gravity through their effect  on the growth of structure. 
A peculiar velocity survey must discover, classify, and get distances of SNe~Ia, as well as obtain redshifts of their host galaxies.
Choices in survey design impact the precision obtained on the growth index $\gamma$.  
No other probe of growth of structure or tracer of peculiar velocity can provide comparable precision on  $\gamma$ in the next decade.
SN~Ia peculiar velocity surveys are of the highest scientific
interest and we encourage
the community to develop aggressive surveys in the pursuit of testing  General Relativity and probing gravity.

@@

Each supernova is assigned a magnitude based on the distance modulus of its cosmological
redshift plus a random term drawn from a Normal distribution with standard deviation $\sigma_M$, which for simplicity is the same for all supernovae and captures both
intrinsic magnitude dispersion and measurement uncertainty.  No other corrections are applied to the observed magnitude.  Dipole effects of the heliocentric
motion with respect to the CMB and of the galaxies with respect to the CMB  are ignored: referring to
\citet{2011ApJ...741...67D}, the effects in Eq.~18 are not included.  Including these effects are straightforward but is computationally expensive
and does not affect our conclusions.
